Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-02-05 article/01_Introduction.Rmd
restructuring the introduction

This article employs some of the methods that arose within archaeology's contemplation of *Cultural transmission theory* to trace the development of four major Bronze Age burial traditions in Europe. The cultural-historical evolution of archaeological cultures and ethnic groups during the European Bronze Age is complex, and its reconstruction an ongoing endeavour of archaeology. Within this history of constant emergence, displacement and demise many different forms of burial traditions were practised. Most of them can be distinguished by the criteria *construction of the grave* and *treatment of the body*: Bronze Age burials are either flat graves or part of a burial mound, and they contain either untreated or cremated human remains.

@@

How can major misconceptions like this arise and persist in European archaeology? Beyond missing data due to preservation and research history, another reason might be a lack of unified data repositories and universally acknowledged methods to measure the spatial and temporal distribution of cultural phenomena. Instead of directly querying archives with quantitative information, the typical process for the creation of large-scale syntheses is still to manually excerpt the literature of all relevant, local research traditions. The present article, in contrast, explores the former strategy. The first major question therefore is: *Does openly available, bulk radiocarbon data allow reconstructing the main trends within the distribution of burial rites in Bronze Age Europe?* This approach could set a precedent for a more reproducible and statistically sound research practice. If we could answer this first question with *yes*, then the quantitative comparison of macro regional trends could pave the way for an analysis of interaction intensity on a long-term continental scale: *Do time series constructed from radiocarbon data enable meaningful observations concerning cultural and spatial distance?*

To answer these questions, a theoretical framework is required that permits the comparison of human behaviour across spatial, temporal and cultural boundaries: The assumption that the cremation aspect of Early Bronze Age burial rituals in Britain could be related to the same aspect in Hungary hundreds of years later is not self-evident. *Cultural evolution theory* provides such a framework [@creanzaCulturalEvolutionaryTheory2017]. The spread of ideas in human populations over space and time is one of it's major topics: *Cultural transmission*.

>>>>

The cultural-historical evolution of archaeological cultures and ethnic groups during the European Bronze Age is complex, and its reconstruction an ongoing endeavour of archaeology. Within this history of constant emergence, displacement and demise many different forms of burial traditions were practised. Most of them can be distinguished by the criteria *construction of the grave* and *treatment of the body*: Bronze Age burials are either flat graves or part of a burial mound, and they contain either untreated or cremated human remains.

@@

How can major misconceptions like this arise and persist in European archaeology? Beyond missing data due to preservation and research history, another reason might be a lack of unified data repositories and universally acknowledged methods to measure the spatial and temporal distribution of cultural phenomena. Instead of directly querying archives with quantitative information, the typical process for the creation of large-scale syntheses is still to manually excerpt the literature of all relevant, local research traditions. The present article, in contrast, explores the former strategy. The first major question therefore is: *Can the main trends in the distribution of burial rites in Bronze Age Europe be detected in openly available bulk radiocarbon data?* This approach could set a precedent for a more reproducible and statistically sound research practice. If we could answer this first question with *yes*, then the quantitative comparison of macro regional trends could pave the way for an analysis of interaction intensity on a long-term continental scale: *Do time series constructed from radiocarbon data enable meaningful observations concerning cultural and spatial distance?*

To answer these questions, a theoretical framework is required that permits the comparison of human behaviour across spatial, temporal and cultural boundaries: The assumption that the cremation aspect of Early Bronze Age burial rituals in Britain could be related to the same aspect in Hungary hundreds of years later is not self-evident. *Cultural evolution theory* provides such a framework [@creanzaCulturalEvolutionaryTheory2017]. The spread of ideas in human populations over space and time is one of its major topics: *Cultural transmission*.
